---
title: "The COVID-19 Pandemic Impacted Stress Differentially in People With Lupus by Social Isolation"
author: "Chenyang Yuan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
---

```{r, set-chunk-opts, echo = FALSE, eval = TRUE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r, load-package, echo = FALSE}
library(kableExtra)
```

*__This report is part of a collaborative project I am working on with Kirk Easley.__*

Many people with systemic lupus erythematosus (SLE), particularly those from communities of color in the U.S., face substantial psychosocial and socioeconomic stressors. The COVID-19 pandemic has significantly impacted mental health. We evaluated the pandemic's effect on stress in people with lupus by examining social isolation.

Georgians Organized Against Lupus (GOAL) is a Centers for Disease Control and Prevention-supported, population-based cohort of validated SLE patients in Atlanta who completed annual surveys across multiple domains. These surveys were conducted over four periods:

* pre-pandemic (baseline): 08/20/2017 -- 01/31/2020

* early-pandemic: 02/01/2020 -- 09/18/2021

* late-pandemic: 09/19/2021 -- 05/11/2023

* post-pandemic: 05/12/2023 -- 10/15/2024

The early- and late- pandemic periods correspond to the U.S. COVID-19 Public Health Emergency declaration and are divided in half. At baseline, we analyzed sociodemographic data and validated measures of lupus characteristics, as well as physical and psychological parameters, including PROMIS social isolation.

We used a mixed model for repeated measures to characterize time trends for the outcome (Perceived Stress Score, PSS). The model was further expanded with stratification to examine the potential effect modification of baseline social isolation on longitudinal PSS trends across these periods.

```{r, specify-path}
here::i_am('report.Rmd')
```

# Descriptive Analysis

The following table displays the baseline demographic characteristics of the patient cohort, stratified into three social isolation groups. Patients experiencing social isolation tended to be younger, have lower socioeconomic status, and exhibit higher levels of disease activity and damage. Baseline stress levels were also higher among those with elevated social isolation.

```{r, table1}
table_one <- readRDS(file = here::here('output/table_one.rds'))
table_one
```

* SLAQ, Systemic Lupus Activity Questionnaire

* BILD, Brief Index of Lupus Damage

# Regression Analysis

We fitted a random intercept model to characterize trends in PSS. Social isolation and time period were the main predictors, with adjustments for covariates including gender, age, SLAQ score, BILD score, poverty status, and insurance type. All covariates were included as categorical variables.

## Graphical analysis

The following figure displays longitudinal trends in PSS across the study periods. Patients with elevated social isolation experienced a reduction in stress at the onset of the pandemic and, over time, did not return to pre-pandemic levels. In contrast, those with typical levels of social isolation showed no significant changes in stress over time.

```{r, plot-data}
#| fig.align = "center",
#| out.width = "600px"
knitr::include_graphics(
  here::here('output/model.png')
)
```

## Regression table

The regression table below presents the regression coefficients and associated p-values for each level of the categorical predictors.

```{r, fit-primary-model}
regression_table <- readRDS(file = here::here('output/regression_table.rds'))
table_row_names <- readRDS(file = here::here('output/table_row_names.rds'))
rownames(regression_table) <- table_row_names$regression
kable(regression_table, digits = c(2, 2, 4), format = "html") %>% kable_styling(full_width = T)
```

## ANOVA table

The ANOVA table below displays the Type-III ANOVA results, highlighting the significance levels of all predictors.

```{r, secondary-model}
anova_table <- readRDS(file = here::here('output/anova_table.rds'))
rownames(anova_table) <- table_row_names$anova
kable(anova_table, digits = c(2, 2, 4), format = "html") %>% kable_styling(full_width = T)
```

# Conclusions

Though this may be unexpected, improvement in stress with the pandemic in those with SLE may have indicated how severe stress had been for these individuals before the pandemic. The impact of social isolation on stress may be mediated through stigmatization and poor functional status. The pandemic may have reduced these factors by bringing people with SLE closer to the general population as the public isolated themselves, wore masks, and implemented occupational and other financial support measures. Improvement in stress was seen most in those with higher levels of social isolation, which may be subgroups to study further and target interventions. Additional research is needed into other related causal pathways, mitigating factors, biological mechanisms, and whether these improvements are durable.

